---
title: "Messy Data Activity"
author: "Dr. Luis Malpica, modified from a script by Dr. Brett Favaro"
date: "February, 2020"
output:
  html_document:
    theme: united
    toc: yes
  pdf_document:
    toc: yes
editor_options: 
  chunk_output_type: console
---

# Introduction
You will look at one table with messy data, you need to clean it up to be able to work with it in R.

Collaboration and teamwork is encouraged!

This dataset contains the following variables:
# netID: A unique identifier for the sampling event (more than one can share)
# fishID: A unique identifier for the individual fish (can't be duplicated)
# species: Species name
# length: Total length (inches to nearest 0.1)
# weight: Wet weight (grams to nearest 0.1)
# year: Year of capture

Good luck!!

# Install required packages
```{r, message=FALSE}
# If you don't yet have tidyverse run:
# install.packages("tidyverse")
library(tidyverse)
```


# STEP 1: Did it load correctly?
```{r}


```


# STEP 2: Check data types
```{r}


```


# STEP 3: Check for missing or impossible values
```{r}


```


# STEP 4: Check for typos and broken factors
```{r}


```



